A heterogeneous landscape of selection and epistasis in protein-coding genes revealed by two-locus statistics

### 2022-03-25 manuscript/paper.tex
getting closer..

\usepackage[margin=1in]{geometry}
\usepackage{url} 
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}

@@

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting allele frequency trajectories and creating correlations
between pairs of mutations. A number of recent studies have used patterns of
linkage disequilibrium (LD) between selected variants to test for selective
interference and epistatic interactions, with some disagreement over
interpreting observations from data. Interpretation is hindered by the relative
lack of analytic or even numerical expectations for patterns of variation
between pairs of loci under the combined effects of selection, dominance,
epistasis, and demography. Here, I develop a numerical approach to compute the
expected two-locus sampling distribution under diploid selection with arbitrary
epistasis and dominance, recombination, and variable population size. I use
this to explore how epistasis and dominance affect expected signed LD,
including for non-steady-state demography relevant to human populations.
\comment{I then use whole-genome sequencing data from humans to assess how well we
can differentiate modes of selective interactions in practice.}
\comment{this needs updating}
I find that
positive LD between missense mutations within genes is driven by strong
positive allele-frequency correlations between pairs of mutations that fall
within the same conserved domain, pointing to compensatory mutations or
antagonistic epistasis as the prevailing mode of interaction within but not
outside of conserved genic elements. The heterogeneous landscape of both
mutational fitness effects and selective interactions within protein-coding
genes calls for more refined inferences of the joint distribution of fitness
and interactive effects, and the methods presented here should prove useful in
that pursuit.

@@

Most new mutations that affect fitness are deleterious and tend to be
eliminated from a population
\textbf{Citations that most mutations are negatively selected}.
The amount of time that a
deleterious mutation segregates depends on the strength of selection against
genomes that carry it, with very damaging mutations kept at low frequencies and
purged relatively rapidly. In the time between mutation and fixation or loss,
selected variants, both beneficial and damaging, can dramatically impact
patterns of variation in nearby linked regions \textbf{MSH, Charlesworth, Stephan/Kim}.
This distortion away from neutral expectations is now empirically documented using
sequencing data from an ever-growing set of study systems, but questions remain
about the primary mode of interactions between multiple linked variants and
their joint effects on genome-wide patterns of diversity.

In their foundational paper, \citet{Hill1966-gv} recognized that linked
selected variants reciprocally impede the efficacy of selection at each locus,
a process known as selective interference. Linked selection reduces the
fixation probability of advantageous mutations and increases that of
deleterious mutations compared to expectations under single-locus models
\citep{Birky1988-jm}. Allele frequency dynamics and correlations of linked
selected variants are also predicted to deviate from expectations without
interference. Under a multiplicative fitness model, where the fitness reduction
of a genome carrying multiple deleterious variants is equal to the product of
the fitness reduction of each mutation independently, we expect to see net
linkage disequilibrium (LD) equal to zero for unlinked sites
\citep{Kondrashov1995-va}. But for linked loci, those mutations are expected to
segregate on different haplotypes more often than together, leading to
negative, or repulsion, LD, although the extent of LD depends non-trivially on
the strength of selection and probability of recombination separating loci
\citep{Hill1966-gv, McVean2000-ox}.

Non-additive effects, including dominance (i.e., interactions \emph{within} a
locus) and epistasis (interactions \emph{between} loci), further complicate our
evolutionary models. A large fraction of nonsynonymous coding mutations are
thought to be at least partially recessive \citep{Agrawal2011-wb,
Huber2018-cp}, with average levels of dominance correlating with strength of
selection \citep{Kacser1981-nc}, and dominance plays an important role in
shaping expected equilibrium allele frequencies and the mutation load of
strongly damaging disease mutations \citep{Clark1998-kq}. On the other hand,
epistasis differentially impacts the deleterious load in asexually and sexually
reproducing organisms \citep{Kimura1966-gc, Kondrashov1995-va}, has been
invoked as an explanation for the evolutionary advantage of sex
\citep{Kondrashov1982-sf, Charlesworth1990-kw, Barton1998-lu}, and can drive
incompatibilities that lead to postzygotic isolation during the process of
speciation \citep{Turelli2000-kz}. Within populations, epistasis is known to
cause signed LD to deviate dramatically from zero \citep{Charlesworth1990-kw,
Kondrashov1995-va}. However, despite appreciation of the effect of dominance on
linked variation \citep{Turelli2000-kz, Zhao2016-bb} and the evolutionary
importance of epistatic interactions, we currently lack models for predicting
correlation patterns between linked mutations under general selection models.

In this paper, I develop a numerical approach to solve for the two-locus
sampling distribution under a general diploid selection model with variable
recombination and single-population size history. I use this model to describe
how epistasis and dominance shape expected patterns of signed LD, under both
steady-state and non-equilibrium demography, that have been used to test for
interference and epistasis in population genomic data. I then turn to human
sequencing data and compare patterns of LD for synonymous, missense, and
loss-of-function mutations within and between protein-coding genes and
annotated domains. I show that while synonymous and missense variants display
similar slightly positive average LD within genes, for missense mutations this
signal is driven by correlations between pairs of mutations within, but not
between or outside of, protein-coding domains.
This suggests an importance for antagonistic (or ``diminishing-returns'')
epistasis or a prevalence of compensatory
nonsynonymous mutations within conserved elements.
%Finally, I discuss how well
%we can expect to discriminate between modes of selective interactions using
%genome-wide patterns of LD.

@@

The most direct way to test for interactions between linked selected variants
is through deep mutation scanning experiments, in which many distinct mutations
are introduced within a target gene and then organismal fitness or some protein
function is experimentally measured \citep{Romero2009-yi, Bank2015-vq,
Puchta2016-rx, Steinberg2016-is}. For example, using the model system of the
TEM-1 \(\beta\)-lactamase gene in \emph{E. coli}, \citet{Bershtein2006-bk}
found evidence for synergistic epistasis, in which the combined effect of
multiple deleterious mutations on individual fitness was larger than would be
expected from multiplying the independent observed effects of each individual
mutation. The scale of mutation scanning experiments continues to improve
dramatically, promising greater resolution of the fitness landscape in such
model systems that can be compared to evolutionary theory
\citep{Otwinowski2018-zk}.

Directed mutational studies are not possible in most natural populations, and
we must turn to population genetic approaches to infer selective interactions
between observed segregating polymorphisms. Motivated by theory that linked
negatively selected mutations will display negative LD due to interference
\citep{Hill1966-gv}, and that epistasis will drive expected LD away from zero,
a number of recent studies have used patterns of LD within classes of
putatively selected variants to infer modes of selective interactions.
\citet{Callahan2011-ac} observed that pairs of tightly linked nonsynonymous
mutations cluster together more than expected along the same lineages in the
\emph{Drosophilid} species complex, and that those clustered mutations tended
to preserve the charge of the protein and were in positive LD compared to pairs
involving synonymous mutations. From this, they proposed that compensatory
nonsynonymous variants are regularly tolerated and maintained. More recently,
\citet{Taverner2020-lk} replicated this finding across a diverse set of genera,
showing that such epistatic interactions are important for protein evolution.

\citet{Sohail2017-zq} observed negative LD between loss-of-function variants in
protein-coding genes (loss-of-function mutations include stop gains and losses,
frameshifts, and other nonsense mutations) in both human and fruit fly
populations. They interpreted this as evidence for widespread synergistic
epistasis between these mutations, in which the fitness reduction of multiple
mutations is greater than the product of that of each individual mutations
independently. Both \citet{Sandler2021-of} and \citet{Garcia2021-zn} have
recently reevaluated patterns of LD between coding variants in humans, fruit
flies, and \emph{Capsella grandiflora}, and suggested interference and
dominance may instead be driving patterns of LD \citep{Garcia2021-zn} or
questioned whether LD between loss-of-function variants is significantly
different from zero \citep{Sandler2021-of}.

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/two-locus-spectrum/stats_summaries}
    \caption{
        \textbf{Sampling distributions and their summaries.}
        Low-order summaries of sampling distributions are commonly computed
        for allele allele frequencies (A, the site-frequency spectrum, SFS)
        and two-locus haplotype distributions (B, linkage disequilibrium, LD).
        Demographic and selective processes affect both the SFS and LD,
        and observations of nonzero values of Tajima's $D$ or signed LD
        ($\sigma_d^1$) are often taken as evidence for selection or interations
        between loci, respectively.
        (C) The full two-locus haplotype sampling distribution is a three-dimensional
        object, making it difficult to visualize. We can instead visualize conditional
        distributions of the full sampling distribution, e.g. conditioned on observing
        five copies of the $A$ allele at the left locus and five copies of $B$ at the
        right locus \citep[e.g.][]{Hudson2001-sg}.
        (D) $\sigma_d^2$, which is closely related to $r^2$, is expected to decay
        with increasing recombination distance between loci \citep{Ohta1969-ie}.
        Selection distorts the level of squared LD away from neutral expectations.
        (E) $\sigma_d^1$, refered to as signed LD here, is expected to be zero for
        pairs of neutral mutations \citep{Hill1968-vu}. Interference between linked
        selected mutation causes negative LD \citep{Hill1966-gv}, and other
        forms of interactions between selected mutations can cause large negative
        or positive signed LD.
        \comment{selected in blue to match all panels, panel labels, and panel A
        should include neutral expectation}
    }
    \label{fig:TwoLocusFS}
\end{figure}

In the Methods, I expand on the moments system developed in
\citet{Ragsdale2019-nt} to compute the expected sampling distribution
\(\Psi_n\) of two-locus haplotypes under a general model of selective
interactions (Figure~\ref{fig:TwoLocusFS}). This sampling distribution stores the
expected density or observed counts of pairs of biallelic loci with each
possible haplotype frequency configuration in a sample of size \(n\). Below, we
compute expectations for \(\Psi_n\) under varying scenarios of selection and
interaction between pairs of loci. It is not possible in this framework to
include the effects of additional linnked selected mutations, such as background
selection due to many linked variants. Individual-based forward simulations are
still needed for such scenarios (e.g., Figures~\ref{fig:bgs1}--\ref{fig:bgs4}).

In many cases it is simpler to visualize summaries such as the expectation or
variance of \(D\) (e.g., Figure~\ref{fig:TwoLocusFS}D,E) or slices of the
distribution (e.g., Figure~\ref{fig:TwoLocusFS}C) instead of the full
three-dimensional distribution. Here I focus on low-order LD statistics
including \(\E[D]\) and \(\E[D^2]\) and their decay with recombination
distance, as these are statistics that are commonly used to test for
interactions between loci. For pairs of biallelic loci, with alleles labeled
\(A/a\) at the left locus and \(B/b\) at the right locus, \(D=f_{AB}-f_A f_B\)
is the standard covariance measure of LD, where \(f_{AB}\) is the frequency of
haplotypes carrying both \(A\) and \(B\), and \(f_A\) and \(f_B\) are the
marginal allele frequencies of \(A\) and \(B\).

Instead of unnormalized \(\E[D^2]\) and \(\E[D]\) I consider expectations for
\(\sigma_d^2 = \E[D^2]/\E[f_A(1-f_A)f_B(1-f_B)]\) and
\(\sigma_d^1 = \E[D]/\E[f_A(1-f_A)f_B(1-f_B)]\).
Such normalized statistics have two benefits:
first, the mutation rate cancels so that expectations are robust to assumptions
about the per-base mutation rate, and second, analytic expectations for these
quantities are available under neutrality at constant population size for
comparison \citep{Ohta1971-yd}.

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/hill_robertson}
    \caption{
        \textbf{Hill-Robertson interference.}
        Interference between pairs of selected mutations can cause deviations
        from expected patterns of LD under neutrality, without dominance
        or epistatic effects.
        (A) The expected normalized variance of $D$ decreases as the strength
        of negative selection increases.
        (B, C) For tightly linked loci (\(4Nr \lesssim 1\)), interference is most
        noticeable for pairs of mutations with \(s \approx 1/N\) as measured by
        signed LD.
        At larger recombination distances (\(4Nr > 1\)), signed LD is most
        negative for stronger selection coefficients. Dashed lines
        show neutral expectations.
    }
    \label{fig:HillRobertson}
\end{figure}

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/epistasis_prediction}
    \caption{
        \textbf{Additive selection and epistasis.}
        Left panels (A and C) show expectations for the decay of
        \(\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]\) with recombination
        distance, and right panesl (B and D) show expectations for the
        decay of \(\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]\). Dashed lines
        show neutral expectations.
        For both weak selection (\(s=-1/2N\), A and B) and stronger selection
        (\(s=-10/2N\), C and D), antagonistic epistasis (\(\epsilon < 0\))
        gives rise to positive signed LD and increased $\sigma_d^2$ over
        a multiplicative model (\(\epsilon = 0\)),
        and synergistic epistasis (\(\epsilon > 0\)) results in negative signed
        LD beyond that of Hill-Robertson interference and decreased \(\sigma_d^2\).
        With epistasis between pairs of selected mutations, nonzero signed LD
    }
    \label{fig:epistasis}
\end{figure}

For mutations under additive selection (\(h=1/2\)) and no epistasis, we
recover the well known \citet{Hill1966-gv} interference result of negative LD
between selected mutations, which is strongest for pairs of mutations that have
selection coefficients \(\gamma = O(1)\), or \(s \approx 1/2N_e\) (Figure
\ref{fig:epistasis}B).
For strongly deleterious mutations, LD is close to zero even
with tight linkage, as they almost always segregate at low enough frequencies
that they are unlikely to interfere with each other \citep{McVean2000-ox}.

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/dominance_roze}
    \caption{
        \textbf{The effect of dominance on LD.}
        (A, B) The strengths of selection and dominance interact in a nonlinear way
        to shape expected signed LD. For weakly to moderately selected mutations,
        as shown here, signed LD can be large and negative for tightly linked loci
        (e.g., \(\gamma=-5\), \(h=0\), and \(\rho < 1\)). However, this large signed
        LD decays with recombination distance faster under a model of recessivity
        than does signed LD under a model of additive selection and
        epistasis (Figure~\ref{fig:epistasis}).
        (C, D) Recessive strongly deleterious mutations can have positive signed
        LD, as recently shown by \citet{Roze2021-cf}.
        However the dominance threshold at which \(\sigma_d^1\) switches from
        positive to negative depends on the strength of selection, and weakly
        selected mutations can show non-monotonic behavior as $h$ varies.
        \comment{C and D labels misplaced}
    }
    \label{fig:dominance}
\end{figure}

%The effects of non-additive selection within a locus and interference has received
%less attention. \citet{Garcia2020-zn} used large-scale forward simulations to explore
%the effect of dominance on patterns of LD, showing that LD depends on both the
%magnitude of the selection and dominance coefficients in a nonlinear way, but they
%note that expensive large-scale simulations are needed to explore these effects
%as no closed form approach to computing those expectations exist.
%The numerical method presented here provides such a closed form approach, so that expectations of signed LD and other two-locus statistics can be computed more
%efficiently and accurately.
\begin{itemize}
    \item Previous work on dominance and LD between linked loci
        \begin{itemize}
            \item \citep{Garcia2021-zn} - large-scale forward simulations to explore the
                effect of dominance on patterns of LD, showing that LD depends on both 
                the magnitude of the selection and dominance coefficients in a nonlinear way
            \item \citep{Roze2021-cf} -
                analytic expression for LD between pairs of recessive
                strongly deleterious mutations under steady-state demography, showing that
                for $h<0.25$, you can get slightly positive signed LD
        \end{itemize}
    \item confirming findings from \citet{Garcia2021-zn} that LD depends in a
        complicated way on the strength of both dominance and selection
    \item confirmed through discrete simulations and numerical approach that positive LD
        can occur for strong negative selection with small $h$
    \item however, Roze's result seems to hold only for strongly negative selection,
        and the $h=0.25$ threshold appears to be a limiting value for that threshold,
        as weak and moderate selection show strikingly different patterns
    \item While the model derived by \citeauthor{Roze2021-cf} showed signed LD is
        positive when $h<0.25$ and negative when $h>0.25$, this appears to 
\end{itemize}

The combined effect of the strength of selection and dominance on interference
is indeed nontrivial, and some parameter regimes can cause strong negative LD
between pairs of negatively selected variants. For strong selection, pairs of
recessive mutations (\(h<1/2\)) can have large negative signed LD (Figure
\ref{fig:dominance}B), although it decays rapidly with increasing distance between
loci and is roughly zero for \(\rho \gg 1\). For weakly selected mutations with
\(\gamma=-1\), we see that there is no monotonic effect of the level of dominance
on negative LD, with both recessive and dominant pairs of mutations having more
negative LD than the additive case.

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/demog_bottle_expand}
    \caption{
        \textbf{The effects of demography on signed LD.}
        Simulations under two toy models, one with an instantaneous expansion in
        the past and the other with a five-fold reduction and recovery illustrated
        in (A), show the effects of population size changes on expected patterns of LD. 
        For two selection strengths (\(\gamma=-1\) and \(\gamma=-10\)) and three
        cases of interactions (non-epistatic interference (\(\epsilon=0\), B),
        synergistic epistasis (\(\epsilon=1/2\), C), and antagonistic epistasis
        (\(\epsilon=-1/2\), D), each with \(\rho=0\)),
        sudden decreases in population size can cause
        large changes in signed LD, often in the opposite direction than more
        subtle shifts due to instantaneous expansion events.
        (E) Signed LD between pairs of rare alleles (both \(n_A, n_B \leq 4\), with
        sample size \(n=50\)) is more sensitive to population size changes.
        (F) Signed LD between pairs of common alleles (both \(n_A, n_B \geq 5/50\))
        is comparatively more stable over time.
        Additional comparisons, including for dominance models and showing
        \(\sigma_d^2\), are shown in Figures~\ref{fig:toy_sd2}--\ref{fig:toy_dom_sd2}.
        Dashed lines indicate neutral expectations.
        \comment{need panel labels}
    }
    \label{fig:toy}
\end{figure}

>>>>

\usepackage[margin=1in]{geometry}
\usepackage{url} 
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}
\usepackage{xspace}

@@

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting allele frequency trajectories and creating correlations
between pairs of mutations. A number of recent studies have used patterns of
linkage disequilibrium (LD) between selected variants to test for selective
interference and epistatic interactions, with some disagreement over
interpreting observations from data. Interpretation is hindered by a lack of
analytic or even numerical expectations for patterns of variation between pairs
of loci under the combined effects of selection, dominance, epistasis, and
demography. Here, I develop a numerical approach to compute the expected
two-locus sampling distribution under diploid selection with arbitrary
epistasis and dominance, recombination, and variable population size. I use
this to explore how epistasis and dominance affect expected signed LD,
including for non-steady-state demography relevant to human populations. Using
whole-genome sequencing data from humans, I explore genome-wide patterns of LD
within protein-coding genes. I show that positive LD between missense mutations
within genes is driven by strong positive allele-frequency correlations between
pairs of mutations that fall within the same annotated conserved domain,
pointing to compensatory mutations or antagonistic epistasis as the prevailing
mode of interaction within conserved genic elements. LD between missense
mutations is reduced outside of conserved domains, as would expected under
Hill-Robertson interference. The heterogeneous landscape of both mutational
fitness effects and selective interactions within protein-coding genes calls
for more refined inferences of the joint distribution of fitness and
interactive effects, and the methods presented here should prove useful in that
pursuit.

@@

Most new mutations that affect fitness are deleterious and tend to be
eliminated from a population. The amount of time that a deleterious mutation
segregates depends on the strength of selection against genomes that carry it,
with very damaging mutations kept at low frequencies and purged relatively
rapidly. In the time between mutation and fixation or loss, selected variants,
both beneficial and damaging, can dramatically impact patterns of variation in
nearby linked regions \citep[e.g.][]{Smith1974-em,Charlesworth1995-dq,Kim2000-on}.
This distortion away from neutral expectations has been empirically documented
using sequencing data from an ever-growing set of study systems, but questions
remain about the primary mode of interactions between multiple linked variants
and their joint effects on genome-wide patterns of diversity.

In their foundational paper, \citet{Hill1966-gv} recognized that linked
selected variants reciprocally impede the efficacy of selection at each locus,
a process known as selective interference. Linked selection reduces the
fixation probability of advantageous mutations and increases that of
deleterious mutations compared to expectations under single-locus models
\citep{Birky1988-jm}. Allele frequency dynamics and correlations of linked
selected variants are also predicted to deviate from single-locus expectations.
Under a multiplicative fitness model, where the fitness reduction of a genome
carrying multiple deleterious variants is equal to the product of the fitness
reduction of each independent mutation, we expect to see net linkage
disequilibrium (LD) equal to zero for unlinked sites \citep{Kondrashov1995-va}.
But for linked loci, those mutations are expected to segregate on different
haplotypes more often than together, leading to negative, or repulsion, LD,
although the extent of LD depends non-trivially on the strength of selection
and probability of recombination separating loci
\citep{Hill1966-gv,McVean2000-ox}.

Non-additive effects, including dominance (i.e., interactions \emph{within} a
locus) and epistasis (interactions \emph{between} loci), further complicate our
evolutionary models. A large fraction of nonsynonymous coding mutations are
thought to be at least partially recessive \citep{Agrawal2011-wb,
Huber2018-cp}, with average levels of dominance correlating with strength of
selection \citep{Kacser1981-nc}, and dominance plays an important role in
shaping expected equilibrium allele frequencies and the mutation load of
strongly damaging disease mutations \citep{Clark1998-kq}. On the other hand,
epistasis differentially impacts the deleterious load in asexually and sexually
reproducing organisms \citep{Kimura1966-gc,Kondrashov1995-va}, has been invoked
as an explanation for the evolutionary advantage of sex
\citep{Kondrashov1982-sf,Charlesworth1990-kw,Barton1998-lu}, and can drive
incompatibilities that lead to postzygotic isolation during the process of
speciation \citep{Turelli2000-kz}. Within populations, epistasis is known to
cause signed LD to deviate dramatically from zero \citep{Charlesworth1990-kw,
Kondrashov1995-va}. However, despite appreciation of the effect of dominance on
linked variation \citep{Turelli2000-kz, Zhao2016-bb} and the evolutionary
importance of epistatic interactions, we currently lack models for predicting
patterns of correlations between linked mutations under general selection
models.

In this paper, I develop a numerical approach to solve for the two-locus
sampling distribution under a general diploid selection model with variable
recombination and single-population size history. I use this model to describe
how epistasis and dominance shape expected patterns of signed LD, under both
steady-state and non-equilibrium demography, that have been used to test for
interference and epistasis in population genomic data. I then turn to human
sequencing data and compare patterns of LD for synonymous, missense, and
loss-of-function mutations in protein-coding genes and annotated conserved
domains. I show that while synonymous and missense variants display similar
slightly positive average LD within genes, for missense mutations this signal
is driven by correlations between pairs of mutations within, but not between or
outside of, protein-coding domains. This suggests an importance for
antagonistic (or ``diminishing-returns'') epistasis or a prevalence of
compensatory nonsynonymous mutations within conserved elements.

@@

The most direct way to test for interactions between linked selected variants
is through deep mutation scanning experiments, in which many distinct mutations
are introduced within a target gene and then organismal fitness or some protein
function is experimentally measured \citep{Romero2009-yi, Bank2015-vq,
Puchta2016-rx, Steinberg2016-is}. For example, using the model system of the
TEM-1 \(\beta\)-lactamase gene in \emph{E. coli}, \citet{Bershtein2006-bk}
found evidence for synergistic epistasis, in which the combined effect of
multiple deleterious mutations on individual fitness was larger than would be
expected from multiplying the independent observed effects of each individual
mutation. The scale of mutation scanning experiments continues to increase,
promising greater resolution of the fitness landscape in such model systems
that can be compared to evolutionary theory \citep{Otwinowski2018-zk}.

Directed mutational studies are not possible in most natural populations, and
we must turn to population genetic approaches to infer selective interactions
between observed segregating polymorphisms. Motivated by theory that linked
negatively selected mutations will display negative LD due to interference
\citep{Hill1966-gv}, and that epistasis will drive expected LD away from zero,
a number of recent studies have used patterns of LD within classes of
putatively selected variants to infer modes of selective interactions.
\citet{Callahan2011-ac} observed that pairs of tightly linked nonsynonymous
mutations cluster together more than expected along the same lineages in the
\emph{Drosophilid} species complex, and that those clustered mutations tend to
preserve the charge of the protein and were in positive LD compared to pairs of
synonymous mutations at the same distance. From this, they proposed that
compensatory nonsynonymous variants are regularly tolerated and maintained.
More recently, \citet{Taverner2020-lk} replicated this finding across a diverse
set of genera, showing that such epistatic interactions are important for
protein evolution.

\citet{Sohail2017-zq} observed negative LD between loss-of-function variants in
protein-coding genes (loss-of-function mutations include stop gains and losses,
frameshifts, and other nonsense mutations) in both human and fruit fly
populations. This was interpreted as evidence for widespread synergistic
epistasis between these mutations, in which the fitness reduction of multiple
mutations is greater than the product of that of each individual mutations
independently. Both \citet{Sandler2021-of} and \citet{Garcia2021-zn} have
recently reevaluated patterns of LD between coding variants in humans, fruit
flies, and \emph{Capsella grandiflora}, and suggested interference and
dominance may instead be driving patterns of LD \citep{Garcia2021-zn} or
questioned whether LD between loss-of-function variants is significantly
different from zero \citep{Sandler2021-of}.

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/two-locus-spectrum/stats_summaries}
    \caption{
        \textbf{Sampling distributions and their summaries.}
        Low-order summaries of sampling distributions are commonly computed
        for allele frequencies (A, the site-frequency spectrum, SFS)
        and two-locus haplotype distributions (B, linkage disequilibrium, LD).
        Demographic and selective processes affect both the SFS and LD,
        and observations of nonzero values of Tajima's $D$ or signed LD
        ($\sigma_d^1$) are often taken as evidence for selection or interactions
        between loci, respectively.
        (C) The full two-locus haplotype sampling distribution is a three-dimensional
        object, making it difficult to visualize. We can instead visualize conditional
        distributions of the full sampling distribution, e.g. conditioned on observing
        five copies of the $A$ allele at the left locus and five copies of $B$ at the
        right locus \citep[e.g.][]{Hudson2001-sg}.
        (D) $\sigma_d^2$, which is closely related to $r^2$, is expected to decay
        with increasing recombination distance between loci \citep{Ohta1969-ie}.
        Selection distorts the level of squared LD away from neutral expectations.
        (E) $\sigma_d^1$, refered to as signed LD here, is expected to be zero for
        pairs of neutral mutations \citep{Hill1968-vu}. Interference between linked
        selected mutation causes negative signed LD \citep{Hill1966-gv}, and other
        forms of interactions between selected mutations can cause large negative
        or positive signed LD.
    }
    \label{fig:TwoLocusFS}
\end{figure}

In the Methods, I expand on the moments system developed in
\citet{Ragsdale2019-nt} to compute the expected sampling distribution
\(\Psi_n\) (Figure~\ref{fig:TwoLocusFS}) of two-locus haplotypes under a
general model of selective interactions. This sampling distribution stores the
expected density or observed counts of pairs of biallelic loci with each
possible haplotype frequency configuration in a sample of size \(n\). Below, we
compute expectations for \(\Psi_n\) under varying scenarios of selection and
interaction between pairs of loci. It is not possible in this framework to
include the effects of additional linked selected mutations, such as background
selection due to many linked variants, and individual-based forward simulations
are still needed for such scenarios (e.g.,
Figures~\ref{fig:bgs1}--\ref{fig:bgs4}).

In many cases it is simpler to visualize summaries such as the expectation or
variance of \(D\) (Figure~\ref{fig:TwoLocusFS}D, E) or conditional slices of the
distribution (Figure~\ref{fig:TwoLocusFS}C) instead of the full
three-dimensional distribution. Here I focus on low-order LD statistics
including \(\E[D]\) and \(\E[D^2]\) and their decay with recombination
distance, as these are statistics that are commonly used to test for
interactions between loci. For pairs of biallelic loci, with alleles labeled
\(A/a\) at the left locus and \(B/b\) at the right locus, \(D=f_{AB}-f_A f_B\)
is the standard covariance measure of LD, where \(f_{AB}\) is the frequency of
haplotypes carrying both \(A\) and \(B\), and \(f_A\) and \(f_B\) are the
marginal allele frequencies of \(A\) and \(B\).

Instead of unnormalized \(\E[D^2]\) and \(\E[D]\) I consider expectations for
\(\sigma_d^2 = \E[D^2]/\E[f_A(1-f_A)f_B(1-f_B)]\) and \(\sigma_d^1 =
\E[D]/\E[f_A(1-f_A)f_B(1-f_B)]\).  Such normalized statistics have two
benefits: first, the mutation rate cancels so that expectations are robust to
assumptions about the per-base mutation rate, and second, we can compare to
analytic expectations for these quantities under neutrality and constant
population size \citep{Ohta1971-yd}.

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/hill_robertson}
    \caption{
        \textbf{Hill-Robertson interference.}
        Interference between pairs of selected mutations can cause deviations
        from expected patterns of LD under neutrality, without dominance
        or epistatic effects.
        (A) The expected normalized variance of $D$ (\(\sigma_d^2\))
        decreases below neutral expectations
        as the strength of negative selection increases.
        (B, C) For tightly linked loci (\(4Nr \lesssim 1\)), interference is most
        noticeable as measured by signed LD for pairs of mutations with
        \(s \approx 1/N\).
        At larger recombination distances (\(4Nr > 1\)), signed LD is most
        negative for somewhat stronger selection coefficients.
        Dashed lines show neutral expectations.
    }
    \label{fig:HillRobertson}
\end{figure}

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/epistasis_prediction}
    \caption{
        \textbf{Additive selection and epistasis.}
        Left panels (A and C) show expectations for the decay of
        \(\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]\) with recombination
        distance, and right panesl (B and D) show expectations for the
        decay of \(\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]\). Dashed lines
        show neutral expectations.
        For both weak (\(s=-1/2N\), A and B) and moderate
        (\(s=-10/2N\), C and D) selection,
        antagonistic epistasis (\(\epsilon < 0\))
        gives rise to positive signed LD and increased $\sigma_d^2$ over
        a multiplicative model (\(\epsilon = 0\)),
        and synergistic epistasis (\(\epsilon > 0\)) results in negative signed
        LD beyond that of Hill-Robertson interference and decreased \(\sigma_d^2\).
    }
    \label{fig:epistasis}
\end{figure}

For mutations under additive selection (\(h=1/2\)) and no epistasis, we
recover the well known \citet{Hill1966-gv} interference result of negative LD
between selected mutations, which is strongest for pairs of mutations that have
selection coefficients \(\gamma = O(1)\), or \(s \approx 1/2N_e\) (Figure
\ref{fig:HillRobertson}).
For strongly deleterious mutations, LD is close to zero even
with tight linkage, as they almost always segregate at low enough frequencies
that they are unlikely to interfere with each other \citep{McVean2000-ox}.

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/dominance_roze}
    \caption{
        \textbf{The effect of dominance on LD.}
        (A, B) The strengths of selection and dominance interact in a nonlinear way
        to shape expected signed LD. For weakly to moderately selected mutations,
        as shown here, signed LD can be large and negative for tightly linked loci
        (e.g., \(\gamma=-5\), \(h=0\), and \(\rho < 1\)). However, this large signed
        LD decays with recombination distance faster under a model of recessivity
        than does signed LD under a model of additive selection and
        epistasis (Figure~\ref{fig:epistasis}).
        (C) Interference effects are most pronounced for recessive deleterious
        variants.
        (D, E) Recessive strongly deleterious mutations can have positive signed
        LD, as recently shown by \citet{Roze2021-cf}.
        However the dominance threshold at which \(\sigma_d^1\) switches from
        positive to negative depends on the strength of selection, and weakly
        selected mutations can show non-monotonic behavior as $h$ varies.
        Selection parameters of \(sh=-0.1\) imply extremely strong selection
        (\(h=0.05\) results in \(s=-0.2\) and \(\gamma=-400\) for homozygous
        diploids at a single locus). The numerical appraoch for \(\Psi_n\)
        cannot handle such strong selection.
    }
    \label{fig:dominance}
\end{figure}

The effect of non-additive selection on correlations between mutations has
received increased attention recently. For example, \citet{Garcia2021-zn} used
large-scale forward simulations to explore how dominance impacts patterns of
LD, showing that LD depends on the magnitudes of both the selection and
dominance coefficients in a nonlinear way. \citet{Roze2021-cf} found an
analytic expression for LD between pairs of strongly deleterious mutations
under steady-state demography, showing that LD can be either positive or
negative depending on the strength of dominance.

The combined effect of the strength of selection and dominance on interference
is indeed nontrivial (Figure~\ref{fig:dominance}A), as observed by
\citet{Garcia2021-zn}. Some parameter regimes can cause strong negative LD
between pairs of negatively selected variants, with moderately selected
recessive variants having stronger signals of interference than additive
selection (Figure~\ref{fig:dominance}B). Unlike epistatic interactions, signed
LD decays rapidly with increasing distance between loci and is roughly zero for
\(\rho \gg 1\). For weakly selected mutations with \(\gamma=-1\), there is no
monotonic effect of the level of dominance on negative LD, with both recessive
and dominant pairs of mutations having more negative LD than the additive case.

Discrete simulations and the moments approach confirm the result from
\citet{Roze2021-cf}, that positive LD can occur for strong negative selection
and small values of \(h\) (Figure~\ref{fig:dominance}D, E). However, while the
analytic formula in \citet{Roze2021-cf} predicts that LD should be positive for
\(h<0.25\) and negative for \(h>0.25\), this appears to only hold in the limit
of strong selection (compare to Figure 1A in \citet{Roze2021-cf}). For moderate
to moderately strong selection, this threshold of \(h\) can be less than
\(0.25\), and LD is negative for all \(0\leq h \leq 1\) for weakly deleterious
mutations with interference strongest between pairs of recessive mutations
(Figure~\ref{fig:dominance}C).

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/demog_bottle_expand}
    \caption{
        \textbf{The effects of demography on signed LD.}
        Simulations under two toy models, one with an instantaneous expansion in
        the past and the other with a five-fold reduction and recovery illustrated
        in (A), show the effects of population size changes on expected patterns of LD. 
        For two selection strengths (\(\gamma=-1\) and \(\gamma=-10\)) and three
        cases of interactions (non-epistatic interference (\(\epsilon=0\), B),
        synergistic epistasis (\(\epsilon=1/2\), C), and antagonistic epistasis
        (\(\epsilon=-1/2\), D), each with \(\rho=0\)),
        sudden decreases in population size can cause
        large changes in signed LD, often in the opposite direction than more
        subtle shifts due to instantaneous expansion events.
        (E) Signed LD between pairs of rare alleles (both \(n_A, n_B \leq 4\), with
        sample size \(n=50\)) is more sensitive to population size changes.
        (F) Signed LD between pairs of common alleles (both \(n_A, n_B \geq 5/50\))
        is comparatively more stable over time.
        Additional comparisons, including for dominance models and showing
        \(\sigma_d^2\), are shown in Figures~\ref{fig:toy_sd2}--\ref{fig:toy_dom_sd2}.
        Dashed lines indicate neutral expectations.
    }
    \label{fig:toy}
\end{figure}
